The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-06 write-up/covid-writeup.tex
Update covid-writeup.tex

Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.  This means selection bias and measurement error can have paradoxical effects.  Here we consider the ratio estimators for the relative change in the prevalence rate.  We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).  Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.  We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error. If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_1/\bar y_2 = y_1/y_2$).  We wish to express the error $\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}$ in statistical terms.  Using a Taylor series approximation, we show that the error can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_2}{\bar Y_1} &\times \bigg[ \rho_{I_2,Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)  -\rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \bigg] \\
&\times \left[ 1 - \rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.  The magnitude of the error depends on the population-level ratio $\bar Y_2 / \bar Y_1$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate. The second term represents the ``cancellation'' that the naive analyst is hoping will occur.  The analyst's intuition only holds when the difference is equal to zero.  This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all held constant.  In the case of New York, for example, recent weeks it seems plausible that $\Delta$ (i.e., selection bias) has not changed drastically.  Figure XX shows the rate of testing remains constant and the total cases relative to the population is small enough such that $f_1 \equiv f_2$ is reasonable. Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported (as far as this author knows).  We will show how such a setting leads to over-estimation prior to and under-estimation after peak prevalence.  The final component is an adjustment to this differential error based on the statistical error at the first time point.

Figure XXa shows a trajectory of bias under several prevalence and data quality scenarios.  We generate data from an SIR model with infection rate $\beta = XX$, total population = $100,000$.  We estimate the bias under a range of values for the relative sampling fraction $f_1/f_0$ and a sampling fraction of $f = 0.05$.  Figure XXb shows the trajectory of $\bar Y_2 / \bar Y_1$.  We see that in each case, the rate of change in prevalence is overestimated prior to the peak and then underestimated afterwards.  From a decision making perspective, such biases have impacts.  First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.  Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic. What both sides miss is that the direction of bias is non-constant over time.  After we hit the peak, estimates will under-estimate prevalence.  Lawmakers and governors will be put under pressure to relax social distancing measures.  This will drive a sequence of cycle of panic and .

@@

\subsubsection{Estimation of effective reproduction number}
Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis (cites).  We show here how these estimates are also impacted by selection bias and measurement error.  For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates. Luis and Ribeiro (2008) show that under a Poisson likelihood, a simple relation between the trajectory of \emph{true new cases} and the effective reproduction number can be derived (see Appendix XX for discussion).  In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19 (Sanche, 2020).  Using this, a method of moments estimator of the effective reproduction number at time $t$ can be given by
$$
R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
$$
Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively. The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$. Let $\hat R_t = 1 + \frac{1}{\gamma} \log \left( y_t / y_{t-1} \right)$ be the estimated effective reproduction number on day $t$.  Under SRS with the same sample size at each step, this is true for sufficiently large sample size.  However, similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.  We again wish to express the statistical error of $\hat R_t - R_t$ in useful terms.  We can re-arrange the error decomposition from Section~\ref{section:rates} to show that this error is given by
$$
\begin{aligned}
\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \bigg] \\
&\times \left[ 1 - \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \right] \bigg)
\end{aligned}
$$
So here we see the exact same trade-off as with the rates but on the logarithmic scale.  For small values $\log(1+x) = x$ so the biases are similar.  However, the error is no longer scaled by the rate of change in prevalence $\bar Y_2/\bar Y_1$. Instead the error is scaled by the serial interval $\gamma^{-1} = 1/7$.  Moreover, the error depends not on aggregate prevalence $\bar Y_t$ but fraction of new cases $\bar K_t$.

@@

So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population (i.e., United States, Germany, France or Michigan, New York, London in isolation).  Many are interested in cross-population comparisons.  Many pieces in the media as well as academic articles have plots of case-count over time aligned by time of first known case.  Others have claimed such comparisons unfair due to unequal population sizes and instead plotted case-count per million.  In this section, we focus on the statistical issues relevant to such comparisons. We do not make any comments on the multi-causal nature of success/failure in countries.

@@

At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express the error $\bar y_1 - \bar y_2$ in statistical terms as
$$
(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  \right]
$$

>>>>

Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.  This means selection bias and measurement error can have paradoxical effects.  Here we consider the ratio estimators for the relative change in the prevalence rate.  We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).  Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.  We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error. If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_t/\bar y_{t-1} = y_t/y_{t-1}$).  We wish to express the error $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_{t}}{\bar Y_{t-1}}$ in statistical terms.  Using a Taylor series approximation, we show that the error can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.  The magnitude of the error depends on the population-level ratio $\bar Y_{t-1} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate. The second term represents the ``cancellation'' that the naive analyst is hoping will occur.  The analyst's intuition only holds when the difference is equal to zero.  This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all held constant.  In the case of New York, for example, recent weeks it seems plausible that $\Delta$ (i.e., selection bias) has not changed drastically.  Figure XX shows the rate of testing remains constant and the total cases relative to the population is small enough such that $f_t \equiv f_{t-1}$ is reasonable. Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported (as far as this author knows).  We will show how such a setting leads to over-estimation prior to and under-estimation after peak prevalence.  The final component is an adjustment to this differential error based on the statistical error at the first time point.

Figure XXa shows a trajectory of bias under several prevalence and data quality scenarios.  We generate data from an SIR model with infection rate $\beta = XX$, total population = $100,000$.  We estimate the bias under a range of values for the relative sampling fraction $f_1/f_0$ and a sampling fraction of $f = 0.05$.  Figure XXb shows the trajectory of $\bar Y_t / \bar Y_{t-1}$.  We see that in each case, the rate of change in prevalence is overestimated prior to the peak and then underestimated afterwards.  From a decision making perspective, such biases have impacts.  First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.  Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic. What both sides miss is that the direction of bias is non-constant over time.  After we hit the peak, estimates will under-estimate prevalence.  Lawmakers and governors will be put under pressure to relax social distancing measures.  This will drive a sequence of cycle of panic and .

@@

\subsubsection{Estimation of effective reproduction number}
Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis (cites).  We show here how these estimates are also impacted by selection bias and measurement error.  For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates. Luis and Ribeiro (2008) show that under a Poisson likelihood, a simple relation between the trajectory of \emph{true new cases} and the effective reproduction number can be derived (see Appendix XX for discussion).  In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19 (Sanche, 2020).  Using this, a method of moments estimator of the effective reproduction number at time $t$ can be given by
$$
R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
$$
Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively. The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$. Let $\hat R_t = 1 + \frac{1}{\gamma} \log \left( y_t / y_{t-1} \right)$ be the estimated effective reproduction number on day $t$.  Under SRS with the same sample size at each step, this is true for sufficiently large sample size.  However, similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.  We again wish to express the statistical error of $\hat R_t - R_t$ in useful terms.  We can re-arrange the error decomposition from Section~\ref{section:rates} to show that this error is given by
$$
\begin{aligned}
\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \bigg] \\
&\times \left[ 1 - \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \right] \bigg)
\end{aligned}
$$
So here we see the exact same trade-off as with the rates but on the logarithmic scale.  For small values $\log(1+x) = x$ so the biases are similar.  However, the error is no longer scaled by the rate of change in prevalence $\bar Y_2/\bar Y_1$. Instead the error is scaled by the serial interval $\gamma^{-1} = 1/7$.  Moreover, the error depends not on aggregate prevalence $\bar Y_t$ but fraction of new cases $\bar K_t$.

@@

So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population (i.e., United States, Germany, France or Michigan, New York, London in isolation) at a single moment and over time.  Many are interested in cross-population comparisons.  Many pieces in the media as well as academic articles have plots of case-count over time aligned by time of first known case.  Others have claimed such comparisons unfair due to unequal population sizes and instead plotted case-count per million.  In this section, we focus on the statistical issues relevant to such comparisons. We do not make any comments on the multi-causal nature of success/failure in countries.

@@

At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express $\bar y_1 - \bar y_2$ in statistical terms as
$$
(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
$$
The first term is the difference in population prevalence (i.e., the quantity of interest).  If random sampling were performed in each population, $E_{\I} [ \rho_{I_j, Y_j}] = 0$ and so the estimate would be unbiased.  However, the second term represents the complex error that results from selection bias.
The particular test of interest is $H_0: \bar Y_1 = \bar Y_2$.  Under the null, $\sigma_{Y_1} = \sigma_{Y_2}$.  In classical statistics, we would compute a Z-score:
$$
\frac{\bar y_1 - \bar y_2}{ V_{SRS} (\bar y_1 - \bar y_2) } =
\frac{\rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} } \overset{\star}{=}
\sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
$$
Where the last equality holds under $f_1 = f_2$, i.e., same sampling rate relative to population size. This tells us that the exact error in the comparison of sample means, as an estimate of the difference in population means, scales as a function of the difference in data quality $\rho_{I_1, Y_1} - \rho_{I_2, Y_2}$ times a square root of a population adjustment.  For highly unequal population sizes (e.g., $N_1 \gg N_2$), the adjustment is approximated well by the square root of the smaller population size minus one (e.g. $\sqrt{N_2 -1}$).

While problematic, this points to a potential setting in which comparisons are still valid: when sampling fraction~$f_1 = f_2$ are equal and data quality are equal.  The issue is how to assess data quality equivalence among populations. \textcolor{red}{Discuss the measures in the literature.}

\subsection{Case count comparisons}
